# Loops

Drawing from Introduction to Programming Logic[@lynne_ohanlon_introduction_2000]
Start with `for` loop as first function term
“Populate an array from a for loop” as an early example; pg 376 blank table with good early exercise.
Side note about infinite loop, importance of settling bounds
What happens if you don’t tell loop to increase?
Examples of `for` loops in multiple languages (Python, C, R)?
To decide: all theory and then all practice, or theory/practice/theory/practice?
Also writing out result of each iteration otherwise you’ll only get the last result (a common problem)


The nice thing about the language used for the control structures in most programming languages is they are usually self-describing.  We're going to talk about loops now.  Loops are common events in programming scripts.  It is what the computer uses to listen for user-input.  It's what researchers use when they want to apply a statistical method over several data files.

The for loop is probably the most common type of loop.  It executes a chunk of code _for_ a certain number of times. The common structure of most for loops is

```
for variable in a range or collection
  do something
```

Let's see what this looks like in action using our list of cheesy mashed potatoes ingredients

```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

for ingredient in ingredents
  write(ingredient)
```

If we asked a computer to do this using the R language, it would look something like this

```{r loop-ingredients}
ingredients <- c("cheddar cheese", "potatoes", "milk", "salt", "butter")
for(ingredient in ingredients){
  print(ingredient)
}
```

But, we want to practice a method of tracing it without the computer doing the work for us. The base set up is like this.  For the original psuedocode, 
```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

for ingredient in ingredents
  write(ingredient)
```
at the top of the paper, write out the `ingredients` assignment, then right below that create headers for all of the variables used within the for loop code block, and add term `write` and another term `loop iteration`.  As you were rewriting this, you may have noticed the typo `ingredents` in the for loop. The nice thing about using pseudocode is you don't need to worry about why your code isn't running if there are typos!

Your paper should look something like this



```{r loop-ingredients-flextable, echo = FALSE, message=FALSE}
library(flextable)

small_border = fp_border(color="grey", width = 1)

ingredients_text <- c("\ningredients <- cheddar cheese, potatoes, milk, salt, butter\n\n")

cName <- c("   ", "loop iteration", " ","ingredient", "  ", "write(ingredient)", "     ") 

ingredients_trace <- matrix(data= " ", nrow = 4, ncol=length(cName))
colnames(ingredients_trace) <- cName
ft <- flextable(as.data.frame(ingredients_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=ingredients_text, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  add_footer_lines(values = c("\n\n ", "\n\n ", "\n\n ", "\n\n "), top = FALSE) %>%
  border_outer(part="footer", border = small_border ) %>%
  border_inner_h(border = small_border, part = "footer")

ft
```

## Exercise - loop trace

Grab a pen and paper and write out the output of each iteration of the loop:

```
for x in 0 to 5 
  write x
  
```
<details>
<summary>Loop output</summary>
0\
1\
2\
3\
4\
5\
</details>\


## Exercise - populate an array

Let's use a `for` loop to populate an array.

We'll start with an empty array called `table` with 4 rows and columns named A, B, C, D, and E:

| A | B | C | D | E |
|---|---|---|---|---|
| &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  |
| &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  |
| &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  |
| &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  | &nbsp;  |

We want to populate this table with numbers, starting with 1 and increasing sequentially (1, 2, 3, 4, etc.). We want to fill each row completely before moving on to start filling the next row.

Our `for` loop:

__[[adapt programming logic book exercise for loop]]__
    

Grab and pen and paper (or a spreadsheet program like Google Sheets or Excel) and manually fill in the empty cells in the array.

- What is the number in cell A1?
- What is the number in cell B2?
- What is the number in cell D4?

<details>
<summary>Your filled array should look like this</summary>

| A   | B   | C   | D   | E  |
|---  |---  |---  |---  |--- |
| 1   |  2  | 3   |  4  | 5  |
| 6   |  7  | 8   |  9  | 10 |
| 11  | 12  | 13  | 14  | 15 |
| 16  | 17  | 18  | 19  | 20 |

</details>


## Other Types of Loops 

There are other types of loops in addition to the commonly used `for` loop.\ 

The `while` loop uses a condition at the start of the loop to determine, in advance, when the loop will stop. For example, recall the initial for loop:

```
for x in 0 to 5 
  write x
  
```

An equivalent `while` loop would look like:

```


```

Another type of loop is a `do while` loop (check at end) and an `until` loop (check at end).

```

```

```

```

## Caveats

Loops are a frequently encountered concept in programming and come with their own set of common issues.

### Infinite loops

When condition is never false or otherwise has no break

### Overwriting outputs

```
for x in 0 to 5 
  answer = x

write answer
  
```




