# Loops

The nice thing about the terms used for the control structures in most programming languages is they are usually self-describing.  We're going to talk about loops now.  

Loops are common events in programming scripts.  It is what the computer uses to listen for user-input.  It's also what researchers use when they want to apply a statistical method over several data files, or several columns within a data table.

## `for` loops

The `for` loop is probably the most common type of loop.  It executes a chunk of code _for_ a certain number of times. The common structure of most `for` loops is

```
for variable in a collection
  do something
```

Now, a collection can be a list of text items, such as cheesy mashed potato ingredients, but it can also be a range of numbers, like `1 to 5` (which is programming speak for 1 2 3 4 5) or `7 to 13` (which is 7 8 9 10 11 12 13).

Let's see what this looks like in action using a basic list of cheesy mashed potatoes ingredients:

```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

for ingredient in ingredients
  write(ingredient)
```

Before we look at any real world code, we will manually go through some example loops, without the computer doing the work for us. Practicing a trace on tasks that are simple, like writing each ingredient in a collection, will help us build skills for, and comfort with, reading complex code. 

### Activity: Practicing a Trace

_Timing: 10 minutes_\
_You will need: a piece of paper and writing implement_

Let's set up our trace exercise. The example loop is: 

```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

for ingredient in ingredients
  write(ingredient)
```
At the top of your paper, write out the `ingredients` assignment (`ingredients <- cheddar cheese, potatoes, milk, salt, butter`).  

On the line below that we're going to create headers for all of the variables used within the `for` loop code block.  Starting on the left, write the term `loop iteration`, then `ingredient` in the center of the line, then the term `write` on the right of the line. 

Your paper should look something like this:

```{r loop-ingredients-flextable, echo = FALSE, message=FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)

small_border = fp_border(color="gray", width = 1)

ingredients_text <- c("\ningredients <- cheddar cheese, potatoes, milk, salt, butter\n\n")

cName <- c("  ", "loop iteration", " ","ingredient", "   ", "write(ingredient)", "    ") 

ingredients_trace <- matrix(data= " ", nrow = 4, ncol=length(cName))
colnames(ingredients_trace) <- cName
ft <- flextable(as.data.frame(ingredients_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=ingredients_text, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  add_footer_lines(values = c("\n\n ", "\n\n ", "\n\n ", "\n\n "), top = FALSE) %>%
  border_outer(part="footer", border = small_border ) %>%
  border_inner_h(border = small_border, part = "footer")

ft
```

Your paper is formatted so it's easy to see what's going on with variable during each loop iteration.  This is called a trace, and is often used in software development to help debug the program.  It is also a good exercise to help build mental muscles to read code.  

The next step is to fill in the values for each iteration of the loop.

Your paper will now look something like this:

```{r loop-ingredients-flextable-mid, echo = FALSE, message=FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)
library(ftExtra) #https://ardata-fr.github.io/flextable-book/extensions.html#ftextra

small_border = fp_border(color="gray", width = 1)

ingredients_text <- c("\ningredients <- cheddar cheese, potatoes, milk, salt, butter\n\n")

cName <- c("   ", "loop iteration", " ","ingredient", "  ", "write(ingredient)", "     ") 

ingredients_trace <- data.frame(one=c("", "", "", "", ""),
                                loop=c("1", "2", "3", "4", "5"),
                                three=c("", "", "", "", ""),
                                ingredient=c("cheddar cheese", "potatoes", "milk", "salt", "butter"),
                                five=c("", "", "", "", ""),
                                write=c("cheddar cheese", "potatoes", "milk *(yes, this is a little redundant)*", "salt *(acknowledge the redunancy)*", "butter *(is there an equivalent to horizontal ditto marks?)*"),
                                seven=c("", "", "", "", "")
                                  )
colnames(ingredients_trace) <- cName

ft <- flextable((ingredients_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=ingredients_text, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  add_footer_lines(values = c("\n\n ", "\n\n"), top = FALSE) %>%
  border_outer(part="footer", border = small_border ) %>%
  border_inner_h(border = small_border, part = "all") %>% 
  colformat_md()

ft
```

Note: It's okay if you didn't completely fill out both the variable `ingredient` and the function `write ingredient` because in this case they are the same.  But, it's important to check what actions are happening to variables when you do a trace.

You've now completed the first trace exercise of a pseudocode example. Congratulations! 

In this workshop, we're focused on the programming logic underlying code, rather than specific syntax, but it can be helpful to look at "real" code doing the same task, to further build out your mental model.

If we asked a computer to do the same task we did, using the R language, it would look something like this:
<hr>

```{r loop-ingredients}
ingredients <- c("cheddar cheese", "potatoes", "milk", "salt", "butter")
for(ingredient in ingredients){
  print(ingredient)
}
```
<hr>

### Activity - Loop Trace 

_Timing: 10 minutes_\
_You will need: a piece of paper and writing implement_

Now that we've practiced a trace using text, let's practice using number ranges  Remember that to a computer, the range written as `2 to 7` represents numbers 2 3 4 5 6 7.

For this loop trace, you'll write out the output of each iteration of this loop:

```
for x in 0 to 5 
  write x
  write x*2
```

Like in the ingredients trace example, you'll create a table with headers for all of the variables used within the `for` loop code block.  Starting on the left, write the term `loop iteration`, then `write(x)` in the center of the line, then the term `write(x*2)` on the right of the line.

Then, fill out the table for each iteration of `for x in 0 to 5`.

<details>
<summary>Loop Trace</summary>
```{r loop-exercise-1, echo = FALSE, message=FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)

small_border = fp_border(color="gray", width = 1)

exercise1_text <- c("\nRange 0 to 5 is 0 1 2 3 4 5\n\n")

cName <- c("   ", "loop iteration", " ","write(x)", "  ", "write(x*2)", "     ") 

exercise1_trace <- data.frame(one=c("", "", "", "", "",""),
                                loop=c("1", "2", "3", "4", "5","6"),
                                three=c("", "", "", "","", ""),
                                x=c("0", "1", "2", "3", "4", "5"),
                                five=c("", "", "", "", "",""),
                                write=c("0", "2", "4", "6", "8", "10"),
                                seven=c("", "", "", "", "","")
                                  )
colnames(exercise1_trace) <- cName

ft <- flextable((exercise1_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=exercise1_text, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  add_footer_lines(values = c("\n\n ", "\n\n"), top = FALSE) %>%
  border_outer(part="footer", border = small_border ) %>%
  border_inner_h(border = small_border, part = "all")

ft
```
</details>\

By this point, you're probably thinking that doing a manual trace is a bit tedious. And it is! But again - this is a good way to strengthen your mental model of how loops work, and get you comfortable with thinking through what a chunk of code is trying to do.


### Exercise - Trace Nested `for` Loops

_Timing: 15 minutes_
_You will need: a piece of paper and writing implement_

So far, we've tackled one loop at a time. But loops are flexible and can be "nested", or embedded within each other.

Let's look at an example the following matrix, which is created using two `for` loops

```{r loop-exercise-2, echo = FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)

# reference: https://davidgohel.github.io/flextable/reference/bg.html

num_rows = 4
num_col = 3
row_names <- list()
matrix_demo <- matrix(nrow=num_rows, ncol = num_col)
for (i in 1:num_rows){
  row_names[i] <- paste0("Row ",i)
  for (j in 1:num_col){
    matrix_demo[i,j] <- (3*(i-1))+(j*2)
  }# end j for loop
}# end i for loop

col_names <- list()
for (j in 1:(num_col)){
  col_names[j] <- paste0("Column ",j)
}
col_names <- c("Index", col_names)


table_demo <-   data.frame(index = unlist(row_names),
                         j1 = matrix_demo[,1],
                         j2 = matrix_demo[,2],
                         j3 = matrix_demo[,3]
                         )

colnames(table_demo) <- col_names

# print(table_demo)

big_border = fp_border(color="black", width = 2)
small_border = fp_border(color="black", width = 1)

ft <- flextable(table_demo) %>% 
      border_remove() %>%
      border_outer(part="all", border = big_border ) %>%
      border_inner_h(part="all", border = small_border ) %>% 
      border_inner_v(part="all", border = small_border ) %>% 
      align(align = "center", part = "all") %>%
      bg(j="Index", bg="gray90", part="body") %>%
      bg(bg="gray90", part="header") %>%
      vline(j="Index", border = big_border, part = "all")
ft 
```


The pseudocode for how this matrix was created is as follows:

```
num_rows = 4
num_col = 3

matrix <- []

for (i in 1 to num_rows)
  for (j in 1 to num_col)
    matrix[i,j] <- (3*(i-1))+(j*2)
```

Before creating the table to help trace the code, answer the following

- What variables need to be traced?
- What are the ranges that are used?
- What calculations do you need to keep track of?
- Is there anything else?

<details>
<summary>Answers</summary>
- What variables need to be traced?
  - `i`, `j`
  - you might think that you need to keep track of `loop iteration`, but it's redundant because both `i` and `j` start at one. it's enough to keep track of just their values.

- What are the ranges that are used?
  - 1 to 4 and 1 to 3

- What calculations do you need to keep track of?
  - `(3*(i-1))+(j*2)`
- Is there anything else?  
  - With a small enough matrix, it's okay to make an empty one and fill it in  
</details><br>

Label columns and sketch out a matrix to fill in the variables as the loop iterates for the first five lines _(you won't fill in the whole matrix, but only a portion)_.

<details>
<summary>Loop Trace</summary>
```{r loop-exercise-2-trace, echo = FALSE, message=FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)
library(ftExtra) #https://ardata-fr.github.io/flextable-book/extensions.html#ftextra

small_border = fp_border(color="gray", width = 1)

exercise2_text_1 <- c("\nnum_rows = 4 & num_col = 3\n\n")
exercise2_text_2 <- c("\nRange 1 to num_rows is 1 2 3 4\n\n")
exercise2_text_3 <- c("\nRange 1 to num_col is 1 2 3\n\n")

cName <- c(" ", "i", "   ", "j", "    ", "(3*(i-1))+(j*2)", "     ") 

exercise1_trace <- data.frame(one=c("", "", "", "", ""),
                                i=c("1", "1", "1", "2", "2"),
                                three=c("", "", "", "", ""),
                                j=c("1", "2", "3", "1", "2"),
                                five=c("", "", "", "", ""),
                                calc=c("(3*(1-1))+(1*2)", "(3*(1-1))+(2*2)", "(3*(1-1))+(3*2)", "(3*(2-1))+(1*2)", "(3*(2-1))+(2*2)"),
                                seven=c("", "", "", "", "")
                                  )
colnames(exercise1_trace) <- cName

ft <- flextable((exercise1_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=exercise2_text_3, top=TRUE, colwidths = length(cName)) %>%
  add_header_row(values=exercise2_text_2, top=TRUE, colwidths = length(cName)) %>%
  add_header_row(values=exercise2_text_1, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  border_outer(part="footer", border = small_border ) %>%
  bg(bg="gray95", part="header") %>%
  bg(i=4, bg="gray90", part="header") %>%
  bold(i=4, bold = TRUE, part = "header") %>%
  italic(i=4, italic = TRUE, part = "header") %>%
  autofit() %>%
  border_inner_h(border = small_border, part = "all")

ft
```


```{r loop-exercise-2-trace-matrix, echo = FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)

# reference: https://davidgohel.github.io/flextable/reference/bg.html

num_rows = 4
num_col = 3
row_names <- list()
matrix_demo <- matrix(nrow=num_rows, ncol = num_col)
for (i in 1:num_rows){
  row_names[i] <- paste0("i = ",i)
  for (j in 1:num_col){
    if(i<2){
      matrix_demo[i,j] <- (3*(i-1))+(j*2)
    }else if(i<3 && j<3){
      matrix_demo[i,j] <- (3*(i-1))+(j*2)
    }else{
      matrix_demo[i,j] <- ""
    }
    
  }# end j for loop
}# end i for loop

col_names <- list()
for (j in 1:(num_col)){
  col_names[j] <- paste0("j = ",j)
}
col_names <- c("Index", col_names)


table_demo <-   data.frame(index = unlist(row_names),
                         j1 = matrix_demo[,1],
                         j2 = matrix_demo[,2],
                         j3 = matrix_demo[,3]
                         )

colnames(table_demo) <- col_names

# print(table_demo)

big_border = fp_border(color="black", width = 2)
small_border = fp_border(color="black", width = 1)

ft <- flextable(table_demo) %>% 
      border_remove() %>%
      border_outer(part="all", border = big_border ) %>%
      border_inner_h(part="all", border = small_border ) %>% 
      border_inner_v(part="all", border = small_border ) %>% 
      align(align = "center", part = "all") %>%
      bg(j="Index", bg="gray90", part="body") %>%
      bg(bg="gray90", part="header") %>%
      vline(j="Index", border = big_border, part = "all")
ft 
```


</details>\

<hr>

## `while` Loops 

There are other types of loops in addition to the commonly used `for` loop.\ 

The `while` loop uses a condition that the computer checks if it is `TRUE` or `FALSE` at the start of the loop to determine if the code chunk inside the loop will execute. 

```
while (condition is TRUE)
  do something
```
This is different from a `for` loop, because while a `for` loop will execute a predetermined number of times, the `while` loop will execute upon a condition being met.

For this loop to work, you need to set the condition *before* the loop, and change it *in* the loop.

You can write a `for` loop into a `while` loop by using a variable which will increment by one for each time the loop is run.  This usually isn't done, but we want to show something you're already familiar with.

For example, recall the initial for loop:

```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

for ingredient in ingredients
  write(ingredient)
```

An equivalent `while` loop would look like:

```
ingredients <- cheddar cheese, potatoes, milk, salt, butter

i = 1
condition_check = length(ingredients) + 1 
while i < condition_check
  write ingredients[i]
  i <- i + 1 // change for condition check
```

Here's what it look like in R

``` {r while-loop}
ingredients <- c("cheddar cheese", "potatoes", "milk", "salt", "butter")

i <- 1
condition_check = length(ingredients) + 1

while(i < condition_check){
  print(ingredients[i])
  i <- i + 1
}
```

When we identify the variables to keep track of in a `while` loop trace, we need to add one to check if the condition for entering the loop is met.  

```{r while-loop-ingredients-flextable, echo = FALSE, message=FALSE}
library(flextable)
library(officer)
library(shiny)
library(dplyr)

small_border = fp_border(color="gray", width = 1)

ingredients_text <- c("\ningredients <- cheddar cheese, potatoes, milk, salt, butter\n\ncondition_check = length(ingredients) = 6\n\n")

cName <- c(" ", "loop iteration", "  ", "i", "   ",  "ingredients[i]", "    ", "i < condition_check", "     ") 

ingredients_trace <- matrix(data= " ", nrow = 4, ncol=length(cName))
colnames(ingredients_trace) <- cName
ft <- flextable(as.data.frame(ingredients_trace)) %>%
  align(align = "center", part = "all") %>%
  add_header_row(values=ingredients_text, top=TRUE, colwidths = length(cName)) %>%
  border_remove() %>%
  border_outer(part="all", border = small_border ) %>%
  add_footer_lines(values = c("\n\n ", "\n\n ", "\n\n ", "\n\n "), top = FALSE) %>%
  border_outer(part="footer", border = small_border ) %>%
  border_inner_h(border = small_border, part = "footer")

ft
```

Try filling out this table for `condition_check = length(ingredients) + 1` (condition_check = 6). What is the last value of `ingredients[i]` that is printed?

Now let's try a thought exercise:

- What would be the last value of `ingredients[i]` printed if `condition_check = 3`?
- What would be the last value of `ingredients[i]` printed if `condition_check = 8`?

<details>
<summary>Answers</summary>
If `condition_check = 3`, then the loop would stop after the `i = 2` iteration. Once `i = 3`, `i < condition_check` would be `False`, and the loop would not continue. So, the last printed value of `ingredients[i]` would be `ingredients[2]`, which is `potatoes`.

In this case, we would be all out of values in `ingredients` before `condition_check` was `False`, meaning the loop would keep going even if there was nothing to return. Since there are only 5 values in `ingredients`, when we got to `i = 6` and `ingredients[6]`, the code would return `NA` (missing value). The same would happen for `i = 7`. So, our full results would look like:
```
[1] "cheddar cheese"
[1] "potatoes"
[1] "milk"
[1] "salt"
[1] "butter"
[1] NA
[1] NA
```
</details>

<hr>

### Other types of loops

Two other types of loops are a `do while` loop (check at end) and an `until` loop (check at end).

```
do
  something
while (condition is TRUE)
```

The logic in the `do while` loop is while a condition is true, the loop will run.  
As with the `while` loop, the condition needs to be set before the loop, and changed within the loop.

For the `until` loop, it's slightly different. The loop structure is like this

```
do
  something
until (condition is TRUE)
```

The logic is opposite of a `do while` loop.  It will run while the condition is false, and will only exit when the condition is true.

These aren't used often, in fact IMO it's better to just re-write the logic to use a `while` loop. 

### Exercise - While Trace

_Timing: 10 minutes_
_You will need: a piece of paper and writing implement_

Let's practice a `while` combined with a `for` trace 

```
condition_check <- 5
i <- 1
while (i < condition_check):
  for j in 1 to 3:
    write (i*j)+i
```

I'll give you all a few minutes to write out your trace.  Once you're done, 
type the largest number you traced in the write statement, but don't press enter until I give the signal.

<hr>
<hr>
## Caveats

As you've just experienced, loops come with their own set of common issues.  
Since loops are a frequently encountered concept in programming, we'll go over the common problems.

### Infinite loops

What you all just experienced in the previous exercise is called an infinite loop.  Infinite loops happen when the condition check is never false.  In the case
of the above while loop, the check variable was never incremented, so the loop would go through the same process until the programmer interrupts it, the computer fails, or time ends (whichever comes first). 

### Overwriting outputs

```
for x in files(1 to 500) 
  rename_file(x, "test-case")
```

Often when researchers are developing a script, they will use test case to develop their algorithm and work out the bugs.  In this way, if mistakes are made, it's on a small scale and easy to correct.  Most development is the following.

1. Develop code for a single case
2. Test on a few cases
3. Use on all of the cases

If a step is missed, it can be disastrous.  I can attest. I wrote a script to rename about 500 files, and in my hubris of being an awesome coder, forgot to test it on a few cases before using it on all of the files. I forgot to change the single rename `("test-case")` to something that incorporated the loop `("test-case"+x)` and when the loop was finished, wondered why I went from 500 to only one file in the folder.  Luckily, I always backup my data files, so it was an easy mistake to remedy _(and makes for a good story on the importance of data backup)_.

## Language examples: Formatting may vary

Programming languages may have specific formatting for loops. This may mean certain brackets must be used, tab indents are needed, or ranges are specified a certain way.

**Python** requires indentation for code blocks.  Code blocks in different structures must be aligned with the same indentation.

Programming structures such as `for` expect `:` at the end of each statement with the subsequent code block intented by one
```python
for x in range(0, 5):
  print(x)
```
<hr>

R does not require indentation for code blocks; however, indentation is used because it's easier for humans to read!

Instead of `:` **R** makes use of curly brackets `{}` to indicate the main body  of a `for` statement.

```r
for (x in 0:5) {
  print(x)
}
```
<hr>
The following example is **C/C++** syntax.  One big difference between C/C++ and Python or R is C/C++ is a compiled programming language, which must first be compiled into a file (often *.exe, but can be anything) that contains the machine-language instructions to be executed.

**C/C++** syntax looks a lot like R, but requires semicolons at the end of each line to be executed in a code block, while R does not require the semicolon (although R will also run if there are semicolons).

```c

```
<hr>

**PHP** is very similar to C/C++, but is used for web development.

```php

```
<hr>

With **Bash**, you need to be careful of spaces around your operators and parentheses. (cheatsheets https://ss64.com/bash/if.html https://devhints.io/bash#conditionals)

This Bash script example compares numbers

```bash

```

This Bash script example compares text

```bash

```

<hr>
